Eco-evolutionary dynamics of finite populations from first principles

### 2023-04-04 ind_files/06_nontechnical_discussion.tex
speak about the new example

The equations of section \ref{sec_fun_theorems} also imply that for the evolution of a trait to be \emph{truly} neutral in finite populations (in the sense of all $m$ types in a system having equal fixation/extinction probability if we start with an initial state in which every type has frequency $1/m$), it is not sufficient for the trait in question to be neutral with respect to fitness $w$. Instead, we also require the trait to be neutral with respect to turnover rate $\tau$. Indeed, in ecological models, previous work in the `quasi-neutral' regime has shown that in finite, fluctuating populations, the equal growth rate of types is not sufficient to ensure equal fixation probabilities and that there is a slight biasing for types with lower turnover rates, sometimes interpreted as a selection `for longevity'~\citep{lin_features_2012, oliveira_advantage_2017,balasekaran_quasi-neutral_2022}. In models of evolutionary game theory in fluctuating finite populations, individuals with lower death rates have higher fixation probability even when growth rates are equalized~\citep{huang_stochastic_2015, czuppon_fixation_2018}. Similarly, models of cell cycle dynamics find that selection favors cell types that periodically arrest their cell cycle relative to non-arresting cells even when their growth rates are equal~\citep{wodarz_effect_2017}. In the language of the birth-death formalism I develop in this thesis, all of these studies equalize the growth rates $w$ of competing types but allow the turnover to vary (by reducing the birth rate through arresting the cell cycle, for example), thus allowing noise-induced selection for reduced turnover to operate in the system.

@@

Being mindful of noise-induced selection is also important for applied fields that regularly deal with manipulating small populations, such as conservation and population management. For example, when trying to increase the population of a hypothetical desired species in a multi-species community, increasing the birth rate is \emph{not} equivalent to reducing the death rate even though both result in an increase in the Malthusian fitness (growth rate) $w_i$. Decreasing the death rate leads to a decrease in $\tau_i$, which is favored by noise-induced selection, whereas increasing the birth rate leads to an increase in $\tau_i$, which leads to noise-induced selection acting to reduce the abundance of the focal species from the community. If the total community size is small, increasing the birth rate of a species can thus lead to noise-induced selection completely eliminating the focal species from the community despite the fact that we \emph{increased} the growth rate of this species. Indeed, numerical investigations of `burst-death' dynamics in fluctuating viral populations show that increasing Malthusian fitness by boosting the survival rate (\emph{i.e.} reducing the death rate) leads to a greater increase in fixation probability than if an exactly equivalent increase in Malthusian fitness is achieved via increasing the burst rate (\emph{i.e.} increasing the birth rate) of viral particles~\citep{alexander_fixation_2008}.  A recent finite population birth-death model for cancer treatment provides another concrete example of the consequences of the asymmetry between changing birth rates and death rates: Due to the presence of noise-induced selection, the potential of a tumorous growth to adapt to treatments and experience evolutionary rescue depends inversely on the per-capita turnover $\tau_i$ of the constituent cancer cells, with obvious implications for optimal treatment strategies~\citep{raatz_promoting_2023}.

>>>>

The equations of section \ref{sec_fun_theorems} also imply that for the evolution of a trait to be \emph{truly} neutral in finite populations (in the sense of all $m$ types in a system having equal fixation/extinction probability if we start with an initial state in which every type has frequency $1/m$), it is not sufficient for the trait in question to be neutral with respect to fitness $w$. Instead, we also require the trait to be neutral with respect to turnover rate $\tau$. Though this is clear from examining equations \eqref{nD_stochastic_RM} or \eqref{nD_stochastic_Price}, I also illustrate a deviation from neutrality explicitly for the $m=2$ case using a simple resource-competition model in Appendix \ref{App_deviation_from_neutrality}. Indeed, in ecological models, previous work in the `quasi-neutral' regime has shown that in finite, fluctuating populations, the equal growth rate of types is not sufficient to ensure equal fixation probabilities and that there is a slight biasing for types with lower turnover rates, sometimes interpreted as a selection `for longevity'~\citep{lin_features_2012, oliveira_advantage_2017,balasekaran_quasi-neutral_2022}. In models of evolutionary game theory in fluctuating finite populations, individuals with lower death rates have higher fixation probability even when growth rates are equalized~\citep{huang_stochastic_2015, czuppon_fixation_2018}. Similarly, models of cell cycle dynamics find that selection favors cell types that periodically arrest their cell cycle relative to non-arresting cells even when their growth rates are equal~\citep{wodarz_effect_2017}. In the language of the birth-death formalism I develop in this thesis, all of these studies equalize the growth rates $w$ of competing types but allow the turnover to vary (by reducing the birth rate through arresting the cell cycle, for example), thus allowing noise-induced selection for reduced turnover to operate in the system.

@@

Being mindful of noise-induced selection is also important for applied fields that regularly deal with manipulating small populations, such as conservation and population management. For example, when trying to increase the population of a hypothetical desired species in a multi-species community, increasing the birth rate is \emph{not} equivalent to reducing the death rate even though both result in an increase in the Malthusian fitness (growth rate) $w_i$. Decreasing the death rate leads to a decrease in $\tau_i$, which is favored by noise-induced selection, whereas increasing the birth rate leads to an increase in $\tau_i$, which leads to noise-induced selection acting to reduce the abundance of the focal species from the community (See Appendix \ref{App_deviation_from_neutrality} for a simple example that illustrates this). If the total community size is small, increasing the birth rate of a species can thus lead to noise-induced selection completely eliminating the focal species from the community despite the fact that we \emph{increased} the growth rate of this species. Indeed, numerical investigations of `burst-death' dynamics in fluctuating viral populations show that increasing Malthusian fitness by boosting the survival rate (\emph{i.e.} reducing the death rate) leads to a greater increase in fixation probability than if an exactly equivalent increase in Malthusian fitness is achieved via increasing the burst rate (\emph{i.e.} increasing the birth rate) of viral particles~\citep{alexander_fixation_2008}.  A recent finite population birth-death model for cancer treatment provides another concrete example of the consequences of the asymmetry between changing birth rates and death rates: Due to the presence of noise-induced selection, the potential of a tumorous growth to adapt to treatments and experience evolutionary rescue depends inversely on the per-capita turnover $\tau_i$ of the constituent cancer cells, with obvious implications for optimal treatment strategies~\citep{raatz_promoting_2023}.

### 2023-04-04 ind_files/99_Appendix_D_examples.tex
speak about the new example

\section{An example of noise-induced selection causing deviations from neutrality despite equal fitness}

@@

Moving to density space via the change of variables $x_i = n_i/K$, letting $\mathbf{x} = [x_1, x_2]^{\mathrm{T}}$, and comparing terms with equations \eqref{nD_functional_forms_for_replicator}, we can see that the per-capita fitness $w_i$ of each type is:
\begin{equation*}
	w_{1}(\mathbf{x}) = w_{2}(\mathbf{x}) = 1 - x_1 - x_2
\end{equation*}
The two types of birds have the same fitness! This implies that $w_1 = w_2 = \overline{w}$ and the selection term in \eqref{nD_replicator_mutator} vanishes.
However, if we now compute the per-capita turnover rates $\tau_i$  of each type, we see that we have
\begin{align*}
	\tau_{1}(\mathbf{x}) &= 1 + x_1 - x_2\\
	\tau_{2}(\mathbf{x}) &= 1 + x_1 + x_2
\end{align*}
Thus, $\tau_{1} < \tau_{2}$, and from equation \eqref{nD_stochastic_RM}, we thus know that noise-induced selection favors type 1 over type 2.
\myfig{1}{figures/App_stoch_LV_density_plots.png}{\textbf{(A)} Time series and \textbf{(B)} Density estimates for $p$, the fraction of type 1 individuals in the population for various values of $K$, obtained from a direct individual-based simulation of the model  conducted via the Gillespie algorithm. At high $K$, the population conforms to deterministic (infinite population) predictions. The time series are from single realizations. The density plots in panel $(B)$ are estimated from 100 independent realizations, each of which were run for $10^4$ timesteps. All simulations were initialized with $n_1 = n_2 = K/2$. In all simulations, $\mu = 0.05$.}{App_stoch_LV_dens_plots}
Direct Gillespie simulations indeed reveal that for low values of $K$, the fraction of individuals in the population that are of type \textcolor{red}{1} is significantly biased to be greater than 0.5, showing the effect of noise-induced selection (Figure \ref{App_stoch_LV_dens_plots}). This bias disappears for high values of $K$, as expected.

For this model, we can in fact quantitatively derive the effects of noise-induced selection by explicitly calculating each term of equation \eqref{nD_stochastic_RM}. Let $p = x_1/(x_1 + x_2)$ be the frequency of type 1 individuals and let $q = 1-p$. Then, it is easy to check by direct substitution of our functional forms that \eqref{nD_stochastic_RM} becomes
\begin{equation}
	dp =  \frac{2}{K}p^2qdt + \mu\left(1-\frac{1}{KN_K}\right)\left(q-p\right) + q\sqrt{A^{+}_1}dW_t^{(1)} - p\sqrt{A^{+}_2}dW_t^{(2)}
\end{equation}
where $A_i^{+}$ is as defined throughout this thesis and each $W^{i}$ is an independent Wiener process. This equation clearly shows the biasing effect of noise-induced selection in the first term of the RHS. Since $p^2(1-p) > 0$ for $p \in (0,1)$, this term always tends to increase the fraction of type 1 individuals in the population. The second term is a mutational term, and simply reflects the fact that we assumed that $1 \to 2$ and $2 \to 1$ mutations are equally likely (and thus the net direction of mutational effects on frequency depends on the difference between the frequencies of type 1 and type 2 individuals).

>>>>

\section{An example of noise-induced selection causing deviations from neutrality despite equal fitness}\label{App_deviation_from_neutrality}

@@

Moving to density space via the change of variables $x_i = n_i/K$, letting $\mathbf{x} = [x_1, x_2]^{\mathrm{T}}$, and comparing terms with equations \eqref{nD_functional_forms_for_replicator}, we can see that the per-capita fitness $w_i$ of each type is:
\begin{equation*}
	w_{1}(\mathbf{x}) = w_{2}(\mathbf{x}) = 1 - x_1 - x_2
\end{equation*}
The two types of birds have the same fitness! This implies that $w_1 = w_2 = \overline{w}$ and the selection term in \eqref{nD_replicator_mutator} vanishes.
However, if we now compute the per-capita turnover rates $\tau_i$  of each type, we see that we have
\begin{align*}
	\tau_{1}(\mathbf{x}) &= 1 + x_1 - x_2\\
	\tau_{2}(\mathbf{x}) &= 1 + x_1 + x_2
\end{align*}
Thus, $\tau_{1} < \tau_{2}$, and from equation \eqref{nD_stochastic_RM}, we thus know that noise-induced selection favors type 1 over type 2.
\myfig{1}{figures/App_stoch_LV_density_plots.png}{\textbf{(A)} Time series and \textbf{(B)} Density estimates for $p$, the fraction of type 1 individuals in the population for various values of $K$, obtained from a direct individual-based simulation of the model  conducted via the Gillespie algorithm. At high $K$, the population conforms to deterministic (infinite population) predictions. The time series are from single realizations. The density plots in panel $(B)$ are estimated from 100 independent realizations, each of which were run for $10^4$ timesteps. All simulations were initialized with $n_1 = n_2 = K/2$. In all simulations, $\mu = 0.05$.}{App_stoch_LV_dens_plots}
Direct Gillespie simulations indeed reveal that for low values of $K$, the fraction of individuals in the population that are of type 1 is significantly biased to be greater than 0.5, showing the effect of noise-induced selection (Figure \ref{App_stoch_LV_dens_plots}). This bias disappears for high values of $K$, as expected.

For this model, we can in fact quantitatively derive the effects of noise-induced selection by explicitly calculating each term of equation \eqref{nD_stochastic_RM}. Let $p = x_1/(x_1 + x_2)$ be the frequency of type 1 individuals and let $q = 1-p$. Then, it is easy to check by direct substitution of our functional forms that \eqref{nD_stochastic_RM} becomes
\begin{equation}
	dp =  \frac{2}{K}p^2qdt + \mu\left(1-\frac{1}{KN_K}\right)\left(q-p\right) + q\sqrt{A^{+}_1}dW_t^{(1)} - p\sqrt{A^{+}_2}dW_t^{(2)}
\end{equation}
where $A_i^{+}$ is as defined throughout this thesis and each $W^{i}$ is an independent Wiener process. This equation clearly shows the biasing effect of noise-induced selection in the first term of the RHS. Since $p^2(1-p) > 0$ for $p \in (0,1)$, this term always tends to increase the fraction of type 1 individuals in the population. The second term is a mutational term, and simply reflects the fact that we assumed that $1 \to 2$ and $2 \to 1$ mutations are equally likely, and thus, the net effect of mutational effects depends on the difference between the frequencies of type 1 and type 2 individuals and flows towards the type with lower frequency, which is why it is $q-p$ and not $p-q$. Finally, the two $dW_t$ terms are non-directional and vanish upon taking an expectation over the probability space, and therefore have no net contribution other than `blurring out the results' if we look at the dynamics averaged over many realizations.
